---
title: "Prepare gene sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Using gene sets from [ScType][1]

`hitype` is designed to be compatible with [ScType][1]. So the gene sets provided by [ScType][1] can be used directly.

```{r}
library(hitype)

gs <- gs_prepare(
  "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_short.xlsx"
)
gs$gene_sets[[1]]$`Acinar cells`

# gs <- gs_prepare(
#  "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_long.xlsx"
# )
```

## Universal marker format (long table)

Besides the wide `ScType` format below, `gs_prepare()` also accepts the
**universal marker format** — a long table with one row per
(cell type, marker) pair, as used across the author's
[biopipen](https://pwwang.github.io/biopipen/) tools. It is also the
default output of `train_weights()` and `find_markers()`.

The table can be a `data.frame` or a file (`.txt`/`.tsv`, `.csv`, `.xlsx`,
`.rds`, or `.qs`/`.qs2` if the qs2 package is installed) with the
following columns:

- `cell_type` (required): Cell type name.
- `gene` (required): Marker gene.
- `direction` (optional): `positive` or `negative` (aliases `pos`/`neg`/`+`/`-`, case-insensitive). Negative markers penalize a cell type.
- `weight` (optional): Numeric magnitude of the marker weight. Combined with `direction`: a positive marker gets weight `abs(weight)`, a negative marker gets `-abs(weight)`. If `direction` is missing, the signed `weight` is used as-is. If neither is given, the marker gets weight 1. Zero-weight markers are allowed.
- `tissue` (optional): Tissue type. One can pass `tissue_type` to `gs_prepare()` to filter gene sets by tissue type.
- `species`, `cancer` (optional): Reserved filter columns (accepted and ignored by `gs_prepare()`).
- `level` (optional): Cell type level, as in the `ScType` format below.

Column names are matched case-insensitively and by alias
(`celltype`/`type` → `cell_type`; `marker`/`gene_symbol` → `gene`;
`sign` → `direction`; `tissueType` → `tissue`).

```{r}
markers <- data.frame(
    cell_type = rep(c("CD4 T", "CD8 T"), each = 3),
    gene = c("CD3E", "CD4", "IL7R", "CD3E", "CD8A", "CD8B"),
    direction = c("positive", "positive", "negative", "positive", "positive", "positive"),
    weight = c(2.0, 1.5, 0.5, 2.0, 1.0, 1.0)
)
gs <- gs_prepare(markers)
gs$gene_sets[[1]]$`CD4 T`
```

## Preparing your own gene sets (wide ScType format)

You can also prepare your own gene sets in the wide `ScType` format.
The gene sets should be a `data.frame` or a tab-delimited file with the
following columns:

- `tissueType` (optional): Tissue type. One can pass `tissue_type` to `gs_prepare()` to filter gene sets by tissue type.
- `cellName` (required): Cell type name.
- `geneSymbolmore1` (required): Markers for the cell type. Multiple markers should be separated by `,`. One can use a suffix `+` to indicate that the marker is a positive marker, and `-` to indicate that the marker is a negative marker. Multiple `+`'s are allowed to indicate that the marker is a strong positive marker. Multiple `-`'s are allowed to indicate that the marker is a strong negative marker. For example, `CD3E+++` indicates that `CD3E` is a strong positive marker, and `CD14---` indicates that `CD14` is a strong negative marker.
- `geneSymbolmore2` (required): Negative markers for the cell type. This column can be empty strings. This is kept for compatibility with [ScType][1]. No suffixes are allowed in this column. A marker in this column is the same as a marker in the previous column with a suffix `-`.
- `level` (optional): Cell type level. The level of the `cellName`. It must start from 1 and increase by 1. The final cell type consists of one `cellName` from each level.
- `nextLevels`: Indication of possible `cellName`s at the next level. Levels should be separated by `;`. The `cellName` at each level should be separated by `,`.
  - The format is `cellName1,cellName2;cellName3,cellName4;...` for each `cellName` at the current level.
  - For example, for `CD4` at level 1, `nextLevels` `Naive;Activated` indicates that `CD4` should be followed by `Naive` at level 2 and `Activated` at level 3.
  - If the `nextLevels` is also specified for `Naive` at level 2, then the `cellName` at level 3 for `CD4` will be the intersection of `Activated` and the `cellName`s at level 3 for `Naive` limited by `nextLevels` for `Naive`.
  - If a level in `nextLevels` is empty, then the `cellName`s at that level will be all `cellName`s at the next level limited by `nextLevels` of that level.
  - `!` can be used to indicate that the `cellName`s at the next level should be excluded. For example, `!Naive` indicates that `Naive` should be excluded from the `cellName`s at the next level.
  - If `!` is the only character in a level, then all `cellName`s at that level will be excluded.

Any `data.frame` of markers (e.g. the output of `train_weights()`) can
be passed directly to `gs_prepare()`:

```{r}
markers <- data.frame(
    cellName = c("CD4 T", "CD8 T"),
    geneSymbolmore1 = c("CD3E,CD4,IL7R", "CD3E,CD8A,CD8B"),
    geneSymbolmore2 = c("", "")
)
gs <- gs_prepare(markers)
```

[1]: https://github.com/IanevskiAleksandr/sc-type