---
title: "Train marker weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Train marker weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You need to install the suggested packages for this vignette. The default
`method = "glmnet"` requires `glmnet`:

```{r, eval = FALSE}
install.packages("glmnet")
devtools::install_github("pwwang/hitype")
```

Other backends have their own requirements (see `?train_weights`):
`ranger` (method `rf`), `xgboost` (method `xgb`), and `keras` +
`innsight` (method `lrp`). For `lrp` you would additionally run:

```{r, eval = FALSE}
install.packages("keras")
install.packages("innsight")
keras::install_keras()
```

To train the weights for the markers, firstly, we have to have the markers in the original `ScType` format:

```{r}
markers <- data.frame(
    cellName = c(
        "Naive CD4+ T", "CD14+ Mono", "Memory CD4+", "B",
        "CD8+ T", "FCFR3A+ Mono", "NK", "DC", "Platelet"
    ),
    geneSymbolmore1 = c(
        "IL7R,CCR7",  "CD14,LYZ", "IL7R,S100A4", "MS4A1",
        "CD8A", "FCGR3A,MS4A7", "GNLY,NKG7", "FCER1A,CST3", "PPBP"
    ),
    geneSymbolmore2 = rep("", 9)
)

markers
```

Prepare the data for training:

```{r, warning=FALSE, message=FALSE}
pbmc <- pbmc3k.SeuratData::pbmc3k
pbmc <- Seurat::UpdateSeuratObject(pbmc)
pbmc[["percent.mt"]] <- Seurat::PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc <- Seurat::NormalizeData(pbmc)
pbmc <- Seurat::FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
pbmc <- Seurat::ScaleData(pbmc, features = rownames(pbmc))
pbmc <- Seurat::RunPCA(pbmc, features = Seurat::VariableFeatures(object = pbmc))
pbmc <- Seurat::FindNeighbors(pbmc, dims = 1:10)
pbmc <- Seurat::FindClusters(pbmc, resolution = 0.5)
pbmc <- Seurat::RunUMAP(pbmc, dims = 1:10)
new_cluster_ids <- markers$cellName
names(new_cluster_ids) <- levels(pbmc)
pbmc <- Seurat::RenameIdents(pbmc, new_cluster_ids)
```

Train the weights with the recommended sparse logistic regression
(`glmnet`) and 5-fold cross-validation. `cv_folds > 1` stratifies the
training cells by cluster and averages the per-fold weights, making the
estimated weights stable and less prone to overfitting a single split:

```{r, eval=requireNamespace("glmnet", quietly = TRUE)}
library(hitype)

weights <- train_weights(
    path_to_gs = markers,
    exprs = pbmc,
    method = "glmnet",
    cv_folds = 5
)
weights
```

Other methods are available via `method =` (e.g. `method = "lrp"` for
the deep-learning LRP backend, `method = "uniform"` for the equal-weight
baseline). For the `lr`, `glmnet`, and `lrp` backends, `cv_folds` is
used for cross-validation:

```{r, eval=FALSE}
weights_lr <- train_weights(markers, pbmc, method = "lr", cv_folds = 5)
weights_lrp <- train_weights(markers, pbmc, method = "lrp", cv_folds = 5)
```

For a statistically sound evaluation, **do not score the trained weights
on the same cells used for training.** Either hold out a test split of
the training dataset, or transfer the weights to another dataset (below).

Use the trained weights on `ifnb` dataset (cross-dataset transfer):

```{r, eval=requireNamespace("ifnb.SeuratData", quietly = TRUE)}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(dplyr))

ifnb <- ifnb.SeuratData::ifnb
ifnb <- Seurat::UpdateSeuratObject(ifnb)

# Keep only the cells that exist in the pbmc dataset
ifnb <- subset(ifnb, subset = seurat_annotations %in% c(
  "CD14 Mono",
  "DC",
  "CD4 Memory T",
  "CD4 Naive T",
  "CD8 T",
  "B",
  "NK"
))

# ifnb ships raw counts in SeuratData — normalize before scoring
ifnb <- Seurat::NormalizeData(ifnb)

ifnb@meta.data <- ifnb@meta.data %>%
  mutate(seurat_clusters = case_when(
    seurat_annotations == "CD14 Mono" ~ "CD14+ Mono",
    seurat_annotations == "CD4 Memory T" ~ "Memory CD4+",
    seurat_annotations == "CD4 Naive T" ~ "Naive CD4+ T",
    seurat_annotations == "CD8 T" ~ "CD8+ T",
    TRUE ~ seurat_annotations
  ))

gs <- gs_prepare(weights)

scores <- hitype_score(Seurat::GetAssayData(ifnb, layer = "data"), gs,
                       norm = "weight", use_sensitivity = FALSE)

assigned <- hitype_assign(ifnb$seurat_clusters, scores, gs, threshold = 0.01)

summary(assigned)
```
## Finding markers from your own data

`find_markers()` discovers marker genes from a labeled dataset (e.g.
`pbmc$seurat_clusters`) and returns them directly in the database format
consumed by `gs_prepare()`, with dependency-light fold-change (`fc`,
default), `Seurat`, or `presto` backends:

```{r, eval=FALSE}
markers <- find_markers(pbmc, clusters = pbmc$seurat_clusters, method = "fc")
weights <- train_weights(path_to_gs = markers, exprs = pbmc, method = "glmnet")
gs <- gs_prepare(weights)
```

See `?find_markers` for details.
